---
title: "assignment_1, Methods 3, 2026"
date: "2026-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
```

```{r, include = FALSE}
pacman::p_load(tidyverse, readbulk, patchwork, lmerTest, ggplot2, emmeans, dplyr)
```

REMEMBER: In your portfolio, make sure to include code that can reproduce the answers requested in the exercises below (**MAKE A KNITTED VERSION**). If it does not KNIT, it cannot be part of your portfolio

# Exercises and objectives

The objectives of the exercises of this assignment are:\
A) Download and organise the data and model\
B) Fit multilevel models for binomial data\
C) Fit multilevel models for count data

--------------------------------------------------- *Alf* ---------------------------------------------------

## Exercise A - Download and organise the data and model

Go to <https://osf.io/ecxsj/files/> and download the files associated with Experiment 2 (there should be 29).\
The data is associated with Experiment 2 of the article at the following DOI <https://doi.org/10.1016/j.concog.2019.03.007>

1)  Put the data from all subjects into a single data frame\

```{r}
data <- read_bulk(
  directory = "experiment_2/",
  fun = read_csv,
  col_types = cols()
  )
```

2)  Describe the data and construct extra variables from the existing variables

<!-- -->

i.  add a variable to the data frame and call it *correct* (have it be a *logical* variable). Assign a 1 to each row where the subject indicated the correct answer and a 0 to each row where the subject indicated the incorrect answer (**Hint:** the variable *obj.resp* indicates whether the subject answered "even", *e* or "odd", *o*, and the variable *target_type* indicates what was actually presented.

```{r}
data <- data |> 
  mutate(correct = ifelse(target.type == "even" & obj.resp == "e" | target.type == "odd" & obj.resp == "o", 1, 0)) 
```

ii. describe what the following variables in the data frame contain, *trial.type*, *pas*, *trial*, *target.contrast*, *cue*, *task*, *target_type*, *rt.subj*, *rt.obj*, *obj.resp*, *subject* and *correct*. (That means you can ignore the rest of the variables in your description). For each of them, indicate and argue for what `class` they should be classified into, e.g. *factor*, *numeric* etc., and apply that class! (**Don't code variables as ordinal; if you think they are ordinal, just code them as a factor**)

## Exercise B - Fit multilevel models for binomial data

We **only** look at the *experiment* trials (*trial.type*)

```{r}

unique(data$trial.type)

exp_data <- data %>%
  filter(trial.type == "experiment")

exp_data$pas <- factor(exp_data$pas)

data$trial.type <- factor(data$trial.type)
cat("trial.type is now:", class(data$trial.type))

data$pas <- factor(data$pas)
cat("pas is now:", class(data$pas))

data$trial <- as.numeric(data$trial)
cat("trial is now:", class(data$trial))

data$target.contrast <- as.numeric(data$target.contrast)
cat("target.contrast is now:", class(data$target.contrast))

data$cue <- factor(data$cue)
cat("cue is now:", class(data$cue))

data$task <- factor(data$task)
cat("task is now:", class(data$task))

data$target.type <- factor(data$target.type)
cat("target.type is now:", class(data$target.type))

data$rt.subj <- as.numeric(data$rt.subj)
cat("rt.subj is now:", class(data$rt.subj))

data$rt.obj <- as.numeric(data$rt.obj)
cat("rt.obj is now:", class(data$rt.obj))

data$obj.resp <- factor(data$obj.resp)
cat("obj.resp is now:", class(data$obj.resp))

data$subject <- factor(data$subject)
cat("subject is now:", class(data$subject))

data$correct <- as.logical(data$correct)
cat("correct is now:", class(data$correct))
```

1)  Fit a multilevel model that models *correct* as dependent on *task* with a unique intercept for each *subject*

```{r}
model.task <- glmer(
  correct ~ task + (1 | subject),
  data = exp_data,
  family = "binomial"
)

model.null <- glmer(
  correct ~ 1 + (1 | subject),
  data = exp_data,
  family = "binomial" 
)
  
 anova_0 <- anova(model.task, model.null)
 
 anova_0
  
```

i.  does *task* explain performance?\

--------------------------------------------------- *Iva* ---------------------------------------------------

ii. add *pas* as a main effect on top of *task* - what are the consequences of that?

```{r}
model.tp <- glmer(
  correct ~ task + pas + (1 | subject),
  data = exp_data,
  family = "binomial"
)

```

```{r}
# Comparing the task and task + PAS models
likelihood_1 <- anova(model.task, model.tp)

likelihood_1

```

```{r}
# Checking coefficients of the model (for further comparison)
coef(summary(model.task))
coef(summary(model.tp))
```

Adding *pas* has improved the model by a lot, as the AIC dropped from 13636 to 12260. The likelihood ratio test, $\chi^2(3) = 1381.90, p < .001$ confirms it.

The effect of *task*, which was initially significant for singles, became non-significant after adding *pas*. Specifically, the coefficient for the singles condition shrank from a significant *b* = 0.17 to a non-significant *b* = -0.04. This suggests that task and *pas* share explanatory variance, and *pas* is the dominant predictor of accuracy in this experiment.

iii. now fit a multilevel model that models *correct* as dependent on *pas* with a unique intercept for each *subject*

```{r}

model.p <- glmer(
  correct ~ pas + (1 | subject),
  data = exp_data,
  family = "binomial"
)
```

<<<<<<< HEAD

iv. finally, fit a model that models the interaction between *task* and *pas* and their main effects, still with a unique intercept for each *subject*\
=======
iii. finally, fit a model that models the interaction between *task* and *pas* and their main effects, still with a unique intercept for each *subject*\
>>>>>>> 28f07361545d6ee8d4dbe6988a7baa2bce6567a3

```{r}

model.tp.inter <- glmer(
  correct ~ task * pas + (1 | subject),
  data = exp_data,
  family = "binomial"
)

```

<<<<<<< HEAD

v. describe in your own words which model is the best in explaining the variance in accuracy, and based on your chosen model - write a short summary on what this says about accuracy as dependent on *pas* and *task*. You should include at least one informative plot that shows the estimated probability of being correct for each of the levels of *task* and *pas* (and their interaction if you included that in your model).
=======
iv. describe in your own words which model is the best in explaining the variance in accuracy, and based on your chosen model - write a short summary on what this says about accuracy as dependent on *pas* and *task*. You should include at least one informative plot that shows the estimated probability of being correct for each of the levels of *task* and *pas* (and their interaction if you included that in your model).
>>>>>>> 28f07361545d6ee8d4dbe6988a7baa2bce6567a3

```{r}
model.null <- glmer(
  correct ~ 1 + (1 | subject),
  data = exp_data,
  family = "binomial"
)
# Comparing all models (including the null model for reference purposes)
likelihood_all <- anova(model.null, model.task, model.p, model.tp, model.tp.inter)

likelihood_all
```

```{r}

# estimated marginal means
emm <- emmeans(model.tp.inter, ~ pas * task, type = "response")

# data frame for ggplot
preds <- as.data.frame(emm)

preds %>%
  ggplot(aes(x = pas, y = prob, fill = task)) +
  geom_col(position = "dodge", width = 0.6, color = "black", alpha = 0.7) +
  
  geom_errorbar(
    aes(ymin = asymp.LCL, ymax = asymp.UCL),
    position = position_dodge(width = 0.6),
    width = 0.2
  ) +
  
  scale_fill_manual(name = "Task", values = c("#90F1EF", "#ffd6e0", "#ffef9f")) +
  labs(
    title = "Relationship between accuracy and PAS",
    x = "PAS",
    y = "Probability of correct response"
  ) +
  
  theme_light() +
  
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  coord_cartesian(ylim = c(0, 1))
```

Answer to v.

The best model for explaining variance in accuracy was the *pas*-only model (correct \~ pas + (1 \| subject)). The interaction between *task* and *pas* was not significant, $\chi^2(6) = 6.64$, *p* = .355, nor did adding task to the PAS model improve the fit (AIC = 12260 vs. 12257), $\chi^2(2) = 0.47$, *p* = .791. The effect of PAS was highly significant, with higher PAS levels strongly predicting higher accuracy (e.g., PAS4 compared to PAS1: *b* = 2.89, *p* < .001).

Therefore, accuracy depends primarily on PAS, while task type does not significantly explain accuracy once PAS is accounted for. We can visually confirm this pattern in the figure above, which displays the estimated probability of a correct response for each level of PAS and task based on the additive model. The figure shows a steep, positive relationship between PAS and accuracy, while the bars for the three task conditions (pairs, quadruplet, singles) overlap almost entirely within each PAS level. This supports the conclusion that task type does not meaningfully influence accuracy.


--------------------------------------------------- *Florin* ---------------------------------------------------

## Exercise C - Fit multilevel models for count data

We **only** look at the *experiment* trials (*trial.type*)

1)  Initialise a new data frame, `data.count`. *count* should indicate the number of times they categorized their experience as *pas* 1-4 for each *task*. I.e. the data frame would have for subject 1: for task:singles, pas1 was used \# times, pas2 was used \# times, pas3 was used \# times and pas4 was used \# times. You would then do the same for task:pairs and task:quadruplet

```{r}
## you can start from this if you want to, but you can also make your own from scratch
data.count <- data.frame(count = numeric(), 
                         pas = numeric(), ## remember to make this into a factor afterwards
                         task = numeric(), ## and this too
                         subject = numeric()) ## and this too
```

2)  Now fit a multilevel model that models a unique "slope" for *pas* for each *subject*, $Z$, with the interaction between *pas* and *task* and their main effects being modelled on the first level, $X$
    i.  which family should be used?\
    ii. why is a slope for *pas* not really being modelled, and what are we modelling instead?\
    iii. if you get a convergence error, try another algorithm (the default is the *Nelder_Mead*) - try (*bobyqa*) for which the `dfoptim` package is needed. In `glmer`, you can add the following for the `control` argument: `glmerControl(optimizer="bobyqa")` (if you are interested, also have a look at the function `allFit`)

--------------------------------------------------- *Dori* ---------------------------------------------------

iv. when you have a converging fit - fit a model with only the main effects of *pas* and *task*. Compare this with the model that also includes the interaction
v.  indicate which of the two models, you would choose and why\
vi. based on your chosen model - write a short summary on what this says about the distribution of ratings as dependent on *pas* and *task* and their interaction. You should include at least one informative plot that shows the estimated rate of events for each of the levels of *task* and *pas* (and their interaction if you included that in your model)
