---
title: "assignment_1, Methods 3, 2026"
date: "2026-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
```

```{r, include = FALSE}
pacman::p_load(tidyverse, readbulk, patchwork, lmerTest)
```

REMEMBER: In your portfolio, make sure to include code that can reproduce the answers requested in the exercises below (**MAKE A KNITTED VERSION**). If it does not KNIT, it cannot be part of your portfolio

# Exercises and objectives

The objectives of the exercises of this assignment are:\
A) Download and organise the data and model\
B) Fit multilevel models for binomial data\
C) Fit multilevel models for count data

--------------------------------------------------- *Alf* ---------------------------------------------------

## Exercise A - Download and organise the data and model

Go to <https://osf.io/ecxsj/files/> and download the files associated with Experiment 2 (there should be 29).\
The data is associated with Experiment 2 of the article at the following DOI <https://doi.org/10.1016/j.concog.2019.03.007>

1)  Put the data from all subjects into a single data frame\

```{r}
data <- read_bulk(
  directory = "experiment_2/",
  fun = read_csv,
  col_types = cols()
  )
```

2)  Describe the data and construct extra variables from the existing variables

<!-- -->

i.  add a variable to the data frame and call it *correct* (have it be a *logical* variable). Assign a 1 to each row where the subject indicated the correct answer and a 0 to each row where the subject indicated the incorrect answer (**Hint:** the variable *obj.resp* indicates whether the subject answered "even", *e* or "odd", *o*, and the variable *target_type* indicates what was actually presented.

```{r}
data <- data |> 
  mutate(correct = ifelse(target.type == "even" & obj.resp == "e" | target.type == "odd" & obj.resp == "o", TRUE, FALSE)) 
```

ii. describe what the following variables in the data frame contain, *trial.type*, *pas*, *trial*, *target.contrast*, *cue*, *task*, *target_type*, *rt.subj*, *rt.obj*, *obj.resp*, *subject* and *correct*. (That means you can ignore the rest of the variables in your description). For each of them, indicate and argue for what `class` they should be classified into, e.g. *factor*, *numeric* etc., and apply that class! (**Don't code variables as ordinal; if you think they are ordinal, just code them as a factor**)

## Exercise B - Fit multilevel models for binomial data

We **only** look at the *experiment* trials (*trial.type*)

1)  Fit a multilevel model that models *correct* as dependent on *task* with a unique intercept for each *subject*
    i.  does *task* explain performance?\

--------------------------------------------------- *Iva* ---------------------------------------------------

ii. add *pas* as a main effect on top of *task* - what are the consequences of that?\
iii. now fit a multilevel model that models *correct* as dependent on *pas* with a unique intercept for each *subject*
iv. finally, fit a model that models the interaction between *task* and *pas* and their main effects, still with a unique intercept for each *subject*\
v.  describe in your own words which model is the best in explaining the variance in accuracy, and based on your chosen model - write a short summary on what this says about accuracy as dependent on *pas* and *task*. You should include at least one informative plot that shows the estimated probability of being correct for each of the levels of *task* and *pas* (and their interaction if you included that in your model).

--------------------------------------------------- *Florin* ---------------------------------------------------

## Exercise C - Fit multilevel models for count data

We **only** look at the *experiment* trials (*trial.type*)

1)  Initialise a new data frame, `data.count`. *count* should indicate the number of times they categorized their experience as *pas* 1-4 for each *task*. I.e. the data frame would have for subject 1: for task:singles, pas1 was used \# times, pas2 was used \# times, pas3 was used \# times and pas4 was used \# times. You would then do the same for task:pairs and task:quadruplet

2)  Now fit a multilevel model that models a unique "slope" for *pas* for each *subject*, $Z$, with the interaction between *pas* and *task* and their main effects being modelled on the first level, $X$
    i.  which family should be used?\
    ii. why is a slope for *pas* not really being modelled, and what are we modelling instead?\
    iii. if you get a convergence error, try another algorithm (the default is the *Nelder_Mead*) - try (*bobyqa*) for which the `dfoptim` package is needed. In `glmer`, you can add the following for the `control` argument: `glmerControl(optimizer="bobyqa")` (if you are interested, also have a look at the function `allFit`)

--------------------------------------------------- *Dori* ---------------------------------------------------

iv. when you have a converging fit - fit a model with only the main effects of *pas* and *task*. Compare this with the model that also includes the interaction
```{r}
# still Florin's code
data.count <- data %>%
  group_by(subject, task, pas) %>%
  count() %>%
  rename(count = n) %>%
  mutate(pas = factor(pas),
         task = factor(task),
         subject = factor(subject))

mlm <- lme4::glmer(count ~ pas + task + pas*task + (pas | subject), data = data.count, family = poisson)
```


```{r}
# New model with main effects of task and pas
m_wo_int <- lme4::glmer(count ~ pas + task + (pas | subject), data = data.count, family = poisson)

```


```{r}
# New dataframe w/ predicted values for both models for 
# visual inspection and further use 

y_hat_1 <- data.frame(predict(mlm, type ="response"))
y_hat_2 <- data.frame(predict(m_wo_int, type ="response"))
data.pred <- cbind(data.count, y_hat_1, y_hat_2) 
colnames(data.pred) <- c("subject", "task", 
                         "pas", "count", "y_hat_1", "y_hat_2")
```

v.  indicate which of the two models, you would choose and why\

My interpretation of the word "better" will be the following: Which model explains the most variance in the current sample, while also keeping in mind the possibility of overfitting; therefore, if a simpler model (one with fewer parameters) explains only marginally less variance than a better model, I will choose the simpler one.

```{r}
install.packages("MuMIn")
library(MuMIn)

print(r.squaredGLMM(mlm))
print(r.squaredGLMM(m_wo_int))
anova_int_noint = anova(mlm, m_wo_int)
anova_int_noint
```
When comparing the two models, I used the $R^2$ value, and an ANOVA. 
**$R^2$ results:** According to the MuMIn package documentation, 
trigamma-estimate is recommended whenever possible, therefore I'll be 
using that as the final $R^2$ value for both models.

For the interaction model, the marginal $R^2$ value was 0.1575 
(rounded to 4 sig. figures) and the conditional $R^2$ value was 0.9749. 
Since the marginal $R^2$ value shows how much of the variance in the data 
is explained by fixed effects, and the conditional $R^2$ value shows the 
variance explained by the model as a whole, it seems like much of the variance 
is being explained by the random effects; i.e.: the model mostly seems to be 
fitting noise. 

For the non-interaction model, a similar trend can be observed: the marginal 
$R^2$ value was 0.1376, which is around the same as for the interaction model.
This model's conditional $R^2$ was also significantly larger 
than its marginal $R^2$, with a value of 0.9745. Overall, the interaction model 
seems to be slightly better so far. 

It is worth noting here that a difference this large in marginal $R^2$ 
and conditional $R^2$ for both models means that the variance explained 
by the fixed effects is small, while the variance explained by random effects 
is large. If we were to use this model on a previously unseen sample, 
it would likely perform poorly.

However, the interaction model has 22 parameters, while the non-interaction 
model only has 16. Is it worth adding these 6 extra parameters just to 
marginally increase the $R^2$ value? The AIC is what I take a look at next.

**ANOVA results:** I compared the AIC values for both models. The interaction 
model has a lower AIC score, at 3148.4, while the non-interaction model's score 
is 3398.5. The interaction model is clearly better here, since the added 
parameters seem to be compensated for by a much better fit according to the 
much lower AIC score. I therefore choose the interaction model.



vi. based on your chosen model - write a short summary on what this says about the distribution of ratings as dependent on *pas* and *task* and their interaction. You should include at least one informative plot that shows the estimated rate of events for each of the levels of *task* and *pas* (and their interaction if you included that in your model)


```{r}
library(ggplot2)

ggplot(data.pred, aes(x = pas, y = y_hat_1, fill = data.pred$task)) +
  geom_violin(alpha = 0.5) +   scale_fill_viridis_d(option = "plasma") +
  facet_wrap(~ task) +
  labs(x = "pas level", y = "count (y_hat)",
       title = "Counts by condition of pas and task") +
  theme_bw() + theme(legend.position="none")

```
The model shows that ratings are consistently higher for *pas* level 1, 2, 
and 4, while *pas* level 3 shows a consistently lower estimated rate of events.

The model also included an interaction between *pas* and *task*. 
In the pairs and quadruplet tasks, ratings varied in a similar way across 
*pas* levels. Overall, *pas* level 1 has the highest variance, with *pas* 
level 3 having the least variance. This changes for the singles 
level of *task*, where *pas* level 3 stils has the least variance, 
but level 2 no longer has the most variance; level 4 does. 


