---
title: "assignment_1, Methods 3, 2026"
date: "2026-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
```

```{r, include = FALSE}
pacman::p_load(tidyverse, readbulk, patchwork, lmerTest)
```

REMEMBER: In your portfolio, make sure to include code that can reproduce the answers requested in the exercises below (**MAKE A KNITTED VERSION**). If it does not KNIT, it cannot be part of your portfolio

# Exercises and objectives

The objectives of the exercises of this assignment are:\
A) Download and organise the data and model\
B) Fit multilevel models for binomial data\
C) Fit multilevel models for count data

--------------------------------------------------- *Alf* ---------------------------------------------------

## Exercise A - Download and organise the data and model

Go to <https://osf.io/ecxsj/files/> and download the files associated with Experiment 2 (there should be 29).\
The data is associated with Experiment 2 of the article at the following DOI <https://doi.org/10.1016/j.concog.2019.03.007>

1)  Put the data from all subjects into a single data frame\

```{r}
data <- read_bulk(
  directory = "experiment_2/",
  fun = read_csv,
  col_types = cols()
  )
```

2)  Describe the data and construct extra variables from the existing variables

i.  add a variable to the data frame and call it *correct* (have it be a *logical* variable). Assign a 1 to each row where the subject indicated the correct answer and a 0 to each row where the subject indicated the incorrect answer (**Hint:** the variable *obj.resp* indicates whether the subject answered "even", *e* or "odd", *o*, and the variable *target_type* indicates what was actually presented.

**Answer:**
We can do this using the mutate() function from dplyr and the ifelse() function.
```{r}
data <- data |> 
  mutate(correct = ifelse(target.type == "even" & obj.resp == "e" | target.type == "odd" & obj.resp == "o", TRUE, FALSE)) 
```

ii. describe what the following variables in the data frame contain, *trial.type*, *pas*, *trial*, *target.contrast*, *cue*, *task*, *target_type*, *rt.subj*, *rt.obj*, *obj.resp*, *subject* and *correct*. (That means you can ignore the rest of the variables in your description). For each of them, indicate and argue for what `class` they should be classified into, e.g. *factor*, *numeric* etc., and apply that class! (**Don't code variables as ordinal; if you think they are ordinal, just code them as a factor**)

**Answer:**

*trial.type*: Trial type is a nominal variables as it holds two different non-numeric options: staircase and experiment. They should therefore be classified as a factor in R as the variable is nominal.
```{r}
data$trial.type <- factor(data$trial.type)
cat("trial.type is now:", class(data$trial.type))
```
*pas*: PAS is the subjective experience that the person rates. It is ordinal, even though it uses numbers, because it has ordering but without a real zero. We still code it as a factor. 
```{r}
data$pas <- factor(data$pas)
cat("pas is now:", class(data$pas))
```
*trial*: This number is the zero-indexed number. This variable can hold discrete information and holds a true zero and should therefore be a ratio, represented as `numeric` in R. 
```{r}
data$trial <- as.numeric(data$trial)
cat("trial is now:", class(data$trial))
```
*target.contrast*: This is the grey-scale proportion of the target digit. This value represents the difference between background and target meaning that a higher value is a bigger different between target and background, making the target easier to see. That means that the variable is a ratio, being represented as `numeric` in R.
```{r}
data$target.contrast <- as.numeric(data$target.contrast)
cat("target.contrast is now:", class(data$target.contrast))
```
*cue*: This is the number for the cue, meaning that it is not continuous, and meaning that it is a nominal variable.  
```{r}
data$cue <- factor(data$cue)
cat("cue is now:", class(data$cue))
```

*task*: Describes how many number could be shown as a target, being either quadruplet, pairs or singles. This holds an inherent ordering, making it ordinal. We still code it as a factor, though.
```{r}
data$task <- factor(data$task)
cat("task is now:", class(data$task))
```
*target.type*: Describes whether the target was even or odd. This is a nominal variable, because it holds no inherent ordering. 
```{r}
data$target.type <- factor(data$target.type)
cat("target.type is now:", class(data$target.type))
```

*rt.subj*: This is the reaction on the PAS score response in seconds. This makes it a ratio variable, since it has a true zero and continuity.  
```{r}
data$rt.subj <- as.numeric(data$rt.subj)
cat("rt.subj is now:", class(data$rt.subj))
```
*rt.obj*: This is the reaction on the target digit response in seconds. This makes it a ratio variable, since it has a true zero and continuity.  
```{r}
data$rt.obj <- as.numeric(data$rt.obj)
cat("rt.obj is now:", class(data$rt.obj))
```
*obj.resp*: This is the key pressed in response to the target, being either e for even or o for odd. This makes it a nominal variable represented as a `factor` in R.
```{r}
data$obj.resp <- factor(data$obj.resp)
cat("obj.resp is now:", class(data$obj.resp))
```
*subject*: This is the subject number, making it nominal, because it has no inherent order and has no true zero. 
```{r}
data$subject <- factor(data$subject)
cat("subject is now:", class(data$subject))
```
*correct*: This is if the subject answered correct on their trial or not. This makes it logical because it holds either true or false. 
```{r}
data$correct <- as.logical(data$correct)
cat("correct is now:", class(data$correct))
```

## Exercise B - Fit multilevel models for binomial data

We **only** look at the *experiment* trials (*trial.type*)

**Answer:** 
We can make sure of this by filtering for only the experimental trials.
```{r}
exp_data <- data |> filter(trial.type == "experiment") 
```


1)  Fit a multilevel model that models *correct* as dependent on *task* with a unique intercept for each *subject*
**Answer:** 
We can do this using the `glmer()` function from the lme4 library, making sure to model as binomial, since our data is binomial.
```{r}
model.task <- glmer(
  correct ~ task + (1|subject),
  data = exp_data,
  family = "binomial")
```


i.  does *task* explain performance?\

**Answer:** 
We can do this by looking at the coefficients of the model.
```{r}
coef(summary(model.task))

#Creating a null model that we can compare the model.task to
model.null <- glmer(correct ~ 1 + (1|subject),
                    data = exp_data,
                    family = binomial)
print("Comparison between models")
print(anova(model.null, model.task))
```
We can from these coefficients see that *task* does indeed explain performance to a certain degree. We can see that as the task includes more numbers, the amount of correct responses decreases, and with less numbers as target it increases. We can see that the singles coefficient is statistically significant and quadruplet isn't, which suggests that task does partly explain performance to some degree. From the anova comparison we see a high $\chi^2$ for the task model compared to a model with without modelling task, which suggests that task does not make the model explain performance much better.
    
--------------------------------------------------- *Iva* ---------------------------------------------------
    
  ii. add *pas* as a main effect on top of *task* - what are the consequences of that?\
  iii. now fit a multilevel model that models *correct* as dependent on *pas* with a unique intercept for each *subject*
  iv. finally, fit a model that models the interaction between *task* and *pas* and their main effects, still with a unique intercept for each *subject*\
  v.  describe in your own words which model is the best in explaining the variance in accuracy, and based on your chosen model - write a short summary on what this says about accuracy as dependent on *pas* and *task*. You should include at least one informative plot that shows the estimated probability of being correct for each of the levels of *task* and *pas* (and their interaction if you included that in your model).

--------------------------------------------------- *Florin* ---------------------------------------------------

## Exercise C - Fit multilevel models for count data

We **only** look at the *experiment* trials (*trial.type*)

1)  Initialise a new data frame, `data.count`. *count* should indicate the number of times they categorized their experience as *pas* 1-4 for each *task*. I.e. the data frame would have for subject 1: for task:singles, pas1 was used \# times, pas2 was used \# times, pas3 was used \# times and pas4 was used \# times. You would then do the same for task:pairs and task:quadruplet

```{r}
## you can start from this if you want to, but you can also make your own from scratch
data.count <- data.frame(count = numeric(), 
                         pas = numeric(), ## remember to make this into a factor afterwards
                         task = numeric(), ## and this too
                         subject = numeric()) ## and this too
```

2)  Now fit a multilevel model that models a unique "slope" for *pas* for each *subject*, $Z$, with the interaction between *pas* and *task* and their main effects being modelled on the first level, $X$
    i.  which family should be used?\
    ii. why is a slope for *pas* not really being modelled, and what are we modelling instead?\
    iii. if you get a convergence error, try another algorithm (the default is the *Nelder_Mead*) - try (*bobyqa*) for which the `dfoptim` package is needed. In `glmer`, you can add the following for the `control` argument: `glmerControl(optimizer="bobyqa")` (if you are interested, also have a look at the function `allFit`)
    
--------------------------------------------------- *Dori* ---------------------------------------------------
    
  iv. when you have a converging fit - fit a model with only the main effects of *pas* and *task*. Compare this with the model that also includes the interaction
  v.  indicate which of the two models, you would choose and why\
  vi. based on your chosen model - write a short summary on what this says about the distribution of ratings as dependent on *pas* and *task* and their interaction. You should include at least one informative plot that shows the estimated rate of events for each of the levels of *task* and *pas* (and their interaction if you included that in your model)
